---
title: "Assignment-1"
date: "2025-09-23"
author: "Nishant Ahlawat"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Statistics-R-Math4753/Lab3")
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install gt only if it's not already available
if (!requireNamespace("gt", quietly = TRUE)) {
  install.packages("gt")
}
library(gt)

```
## I completed 15/15 tasks
# Task 1
```{r}

grading <- data.frame(
  Group = c("Labs", "Chapter Quizzes", "Assignments", "Projects",
            "In class Quizzes", "Midterm exams", "Final exam"),
  Count = c("16", "9", "4", "2", "TBD", "1", "1"),
  `Weight (each)` = c("0.625%", "0.56%", "3.75%", "3.33%, 6.67%", "-", "20%", "30%"),
  `Total Weight`  = c("10%", "5%", "15%", "10%", "10%", "20%", "30%"),
  check.names = FALSE
)

grading |>
  gt() |>
  tab_header(
    title = md("**Overall summary of the grading for this course**")
  ) |>
  cols_align(align = "left", columns = "Group") |>
  cols_align(align = "center", columns = c("Count", "Weight (each)", "Total Weight")) |>
  tab_options(
    table.font.names = "Courier",         # monospaced vibe
    data_row.padding = px(6),
    table.border.top.width = px(1),
    table.border.bottom.width = px(1)
  ) |>
  tab_source_note(md("_And, true to form, the right column sums to 100%!_")) |>
  tab_source_note(md(
    "There will be no curves in this course. A final grade in the 90s earns an **A**, \
a final grade in the 80s earns a **B**, a final grade in the 60s or 70s earns a **C**, \
a final grade in the 50s earns a **D**, and anything below 50 earns an **F**."
  ))

```

# Task 2
## (a)
```{r}
ddt <- read.csv("DDT.csv")

# Treat MILE as factor
m <- with(ddt, as.numeric(factor(MILE)))

# Coplot LENGTH vs WEIGHT given RIVER*SPECIES
coplot(LENGTH ~ WEIGHT | RIVER * SPECIES,
       data = ddt,
       pch = m,                     # different plotting symbols
       col = m,                     # different colors
       panel = panel.smooth,        # adds smooth line
       xlab = "Weight", ylab = "Length")

legend("topright",
       legend = levels(factor(ddt$MILE)),
       col = 1:length(unique(m)),
       pch = 1:length(unique(m)),
       title = "MILE")
```
## (b)
"The lower left panels correspond to the first RIVER-SPECIES combinations.
They show positive association between LENGTH and WEIGHT: larger fish weigh more.

## (c) what does line A do?
m = with(ddt, as.numeric(factor(MILE)))
"This converts the factor MILE into numeric codes (1, 2, 3, …).
Each unique MILE gets a number that we can use for plotting symbols (pch) and colors."

## (d) What does line B do?
length(unique(m))
"This counts how many unique MILE categories exist.
It ensures we know how many distinct colors/symbols are needed for plotting."

## (e) Why are the top six plots empty?
"The coplot is conditioned on RIVER*SPECIES.
If no fish were caught for some RIVER-SPECIES combinations in the dataset, those panels have no data → they appear empty."

## (f)
```{r}
mean_ddt <- mean(ddt$DDT[ddt$RIVER == "FCM" & ddt$SPECIES == "CCATFISH"])
mean_ddt


```
# Task 3

"All highway bridges in the United States are inspected periodically for structural deficiency by the FHWA. Several variables from the National Bridge Inventory are given below. Classify each as **quantitative** or **qualitative**."

| Variable | Type |
|---|---|
| a. Length of maximum span (feet) | Quantitative |
| b. Number of vehicle lanes | Quantitative |
| c. Toll bridge (yes or no) | Qualitative |
| d. Average daily traffic | Quantitative |
| e. Condition of deck (good, fair, or poor) | Qualitative (ordinal) |
| f. Bypass or detour length (miles) | Quantitative |
| g. Route type (interstate, U.S., state, county, or city) | Qualitative (nominal) |

## Notes

- **Quantitative**: numeric and measurable (continuous or discrete).  
- **Qualitative**: categorical.  
  - **Ordinal**: categories have a natural order (e.g., good > fair > poor).  
  - **Nominal**: categories have no inherent order (e.g., route type).

# Task 4- Random Sampling Designs

## (a) The four random sampling designs are:

1. Simple random sampling  
2. Stratified random sampling  
3. Cluster sampling  
4. Systematic sampling  

---

## (b) **Descriptions of each sampling design**

- **Simple random sampling**:  
  Every experimental unit in the population has an equal chance of being selected. For example, drawing names from a hat or using a random number generator.

- **Stratified random sampling**:  
  The population is divided into groups (called strata) that are internally homogeneous (similar within each group) but different from one another. A random sample is then taken from each stratum, ensuring representation across all strata.

- **Cluster sampling**:  
  The population is divided into large groups (clusters) that mirror the overall population. A random sample of entire clusters is selected, and then all experimental units within the chosen clusters are observed.

- **Systematic sampling**:  
  A starting point is chosen at random, and then every *k*th experimental unit is selected from a list or sequence. For example, surveying every 10th person on a list.

# Task 5
We have data on 223 New Hampshire wells (`MTBE.csv`). The task is to select a random sample of 5 wells and answer follow-up questions.

---

## Random sample of 5 wells

```{r}
# Load libraries
library(knitr)
# Load data
mtbe <- read.csv("MTBE.csv", header = TRUE)

# First six rows (preview of dataset)
kable(head(mtbe), caption = "First six rows of MTBE data")

# Dimensions (rows and columns)
dim(mtbe)

# Select random indices
set.seed(123)   # for reproducibility
ind <- sample(1:223, 5, replace = FALSE)

# Random sample of 5 wells
sample_wells <- mtbe[ind, ]
kable(sample_wells, caption = "Random sample of 5 wells")

## (i) Remove rows with missing values
# Remove rows with NA values
mtbeo <- na.omit(mtbe)

# Show cleaned dataset dimensions
dim(mtbeo)

# Show first few cleaned rows
kable(head(mtbeo), caption = "First rows of MTBE data with NAs removed")

## (ii) Standard deviation of depth for wells with Bedrock aquifer
# Subset wells where Aquifier == "Bedrock"
depth <- mtbeo[mtbeo$Aquifier == "Bedrock", ]$Depth

# Compute standard deviation
sd_depth <- sd(depth, na.rm = TRUE)
sd_depth
```

# Task 6
The U.S. Geological Survey recorded 2,929 aftershocks in the Los Angeles area following a major earthquake. The data (measured on the Richter scale) are in the file `EARTHQUAKE.csv`. We want to:

1. Select a random sample of 30 aftershocks.  
2. Plot the magnitudes as a time series.  
3. Compute the median magnitude for the full dataset.  

---

## Random sample of 30 aftershocks

```{r}
# Load libraries
library(knitr)

# Load dataset
eq <- read.csv("EARTHQUAKE.csv", header = TRUE)

# First rows preview
kable(head(eq), caption = "First six rows of Earthquake data")

# Dataset dimensions
dim(eq)

# Select a random sample of 30 aftershocks
set.seed(123)  # reproducibility
ind <- sample(1:nrow(eq), 30, replace = FALSE)
sample_eq <- eq[ind, ]

# Show sample in a clean table
kable(sample_eq, caption = "Random sample of 30 aftershocks")
```

## (i) Plot magnitudes as a time series
```{r}
# Time series plot of magnitudes
plot(ts(eq$MAG),
     main = "Time Series of Aftershock Magnitudes",
     ylab = "Magnitude (Richter scale)",
     xlab = "Observation Index",
     col = "blue")
```

## (ii) Median magnitude of all aftershocks

```{r}
median_mag <- median(eq$MAG, na.rm = TRUE)
median_mag

```

# Task 7
## (a) What is the data collection method?

The data collection method is a **designed experiment**.  
Engineers collected fish specimens at predetermined river and tributary locations, measuring weight, length, and DDT concentration.

---

## (b) What is the population?

The population is **all fish in the Tennessee River and its tributaries**.  
The 144 fish captured and measured represent a **sample** from this larger population.

---

## (c) Give the names of all the qualitative variables.

The qualitative variables are:  

- **River** (capture location: TR, FC, LC, SC)  
- **Mile** (distance upstream, categorical in this study)  
- **Species** (channel catfish, largemouth bass, smallmouth buffalo)

---

## Notes

- **Qualitative variables** describe categories or labels.  
- **Quantitative variables** (not asked here, but present) include weight (g), length (cm), and DDT concentration (ppm).

# Task 8

## (a) What type of graph is used to describe the data?

A **bar graph** is used to describe the data.

---

## (b) What is the variable measured?

The variable measured is **type of robotic limbs** for each robot.

---

## (c) Which design is used the most?

The **“Legs only”** design (63 robots) is the most common.

---

## (d) Compute class relative frequencies

```{r}
# Frequencies of robot limb categories
freq <- c(15, 8, 63, 20)
RL <- c("None", "Both", "Legs only", "Wheels only")

# Data frame
robots <- data.frame(Category = RL, Frequency = freq)

# Total sample size
n <- sum(freq)

# Relative frequencies
robots$RelativeFreq <- round(robots$Frequency / n, 4)

robots

knitr::kable(robots, digits = 3, caption = "Frequencies and Relative Frequencies of Robot Designs")
```
## (e) Pareto diagram
```{r}
library(qcc)

pareto.chart(freq,
             names = RL,
             main = "Pareto Diagram of Robotic Limb Categories",
             ylab = "Number of Robots")


```
# Task 9

Researchers analyzed 50 Microsoft security bulletins from 2012.  
- Windows: 32  
- Explorer: 6  
- Office: 12  

Categories of security repercussions:  
- Denial of service = 6  
- Information disclosure = 8  
- Remote code execution = 22  
- Spoofing = 3  
- Privilege elevation = 11  

---

## (a) Pie chart of Microsoft products with security issues

```{r}
products <- c("Windows", "Explorer", "Office")
counts <- c(32, 6, 12)

# Pie chart
pie(counts,
    labels = products,
    main = "Microsoft Products with Security Issues (2012)",
    col = c("skyblue", "orange", "lightgreen"))

```

## (b) Pareto diagram of repercussions

```{r}
# Categories of repercussions
issues <- c("Denial of service", "Information disclosure", 
            "Remote code execution", "Spoofing", "Privilege elevation")
freq <- c(6, 8, 22, 3, 11)

# If qcc not installed: install.packages("qcc")
library(qcc)

pareto.chart(freq,
             names = issues,
             main = "Pareto Chart of Security Issue Repercussions",
             ylab = "Number of Bulletins")
```

# Task 10
```{r}
# Load package
library(plotrix)

# Load data
swd <- read.csv("SWDEFECTS.csv", header = TRUE)

# Tabulate defects
tab <- table(swd$defect)
rtab <- tab / sum(tab)

# Relative frequencies
round(rtab, 2)

# 3D pie chart
pie3D(rtab,
      labels = c("OK", "Defective"),
      main = "Pie Chart of Software Defects",
      col = c("lightblue", "red"))

```

# Task 11

## Data

```{r}
old <- c(9.98, 10.12, 9.84, 10.26, 10.05, 10.15, 10.05, 9.80, 10.02, 
         10.29, 10.15, 9.80, 10.03, 10.00, 9.73, 8.05, 9.87, 10.01,
         10.55, 9.55, 9.98, 10.26, 9.95, 8.72, 9.97, 9.70, 8.80, 
         9.87, 8.72, 9.84)

new <- c(9.19, 10.01, 8.82, 9.63, 8.82, 8.65, 10.10, 9.43, 8.51,
         9.70, 10.03, 9.14, 10.09, 9.85, 9.75, 9.60, 9.27, 8.78,
         10.05, 8.83, 9.35, 10.12, 9.39, 9.54, 9.49, 9.48, 9.36,
         9.37, 9.64, 8.68)
```

## (a) Relative frequency histogram – Old Location
```{r}
# Define bins 8.0–10.6 with 9 classes

bins <- seq(8.0, 10.6, length.out = 10)

old_cut <- cut(old, breaks = bins, right = TRUE, include.lowest = TRUE)
old_tab <- table(old_cut)
old_rel <- old_tab / length(old)

barplot(old_rel, space = 0, col = "skyblue",
        main = "Relative Frequency Histogram – Old Location",
        ylab = "Relative Frequency", las = 2)
old_rel

```
## (b) Stem-and-leaf display
```{r}
stem(old)
```
## (c) Histogram (using barplot, not hist)
```{r}
new_cut <- cut(new, breaks = bins, right = TRUE, include.lowest = TRUE)
new_tab <- table(new_cut)
new_rel <- new_tab / length(new)

barplot(new_rel, space = 0, col = "orange",
        main = "Relative Frequency Histogram – New Location",
        ylab = "Relative Frequency", las = 2)
new_rel

```
## (d) Compare Old vs. New

```{r}
barplot(rbind(old_rel, new_rel), beside = TRUE,
        col = c("skyblue", "orange"),
        legend.text = c("Old", "New"),
        main = "Comparison of Old vs. New Voltage Readings",
        ylab = "Relative Frequency", las = 2)

```
## (d) Compare Old vs. New

```{r}
barplot(rbind(old_rel, new_rel), beside = TRUE,
        col = c("skyblue", "orange"),
        legend.text = c("Old", "New"),
        main = "Comparison of Old vs. New Voltage Readings",
        ylab = "Relative Frequency", las = 2)

```
## (e) Mean, Median, Mode

```{r}
mean_old <- mean(old); median_old <- median(old)
mean_new <- mean(new); median_new <- median(new)

# Mode function
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

list(Old = c(Mean = mean_old, Median = median_old, Mode = get_mode(old)),
     New = c(Mean = mean_new, Median = median_new, Mode = get_mode(new)))

```
## (f) Z-score at 10.50 (Old)

```{r}
z_old <- (10.50 - mean_old) / sd(old)
z_old

```
## (g) Z-score at 10.50 (New)

```{r}
z_new <- (10.50 - mean_new) / sd(new)
z_new

```

## (h) Where is 10.50 more likely?

Answer: Compare |z| values: whichever is smaller means 10.50 is more typical there. i.e. OLD LOCATION

## (i-l) Boxplots and Outliers

```{r}
par(mfrow=c(1,2))
boxplot(old, main = "Boxplot – Old Location", col = "skyblue")
boxplot(new, main = "Boxplot – New Location", col = "orange")
par(mfrow=c(1,1))

# Outlier detection using z-scores
outliers_old <- old[abs((old - mean_old)/sd(old)) > 2]
outliers_new <- new[abs((new - mean_new)/sd(new)) > 2]
list(OutliersOld = outliers_old, OutliersNew = outliers_new)


```

## (m) Compare Old vs. New

Answer:

Old readings range lower, with some below the 9.2 threshold.

New readings are more tightly clustered and generally higher.

The new process seems better suited for meeting the quality standard.

# Task 12

Data (micrometers) from *Anti-corrosion Methods and Materials* (Vol. 50, 2003).  
Goal: Give an interval that will **likely contain about 95%** of coated pipe roughness measurements.

```{r}
rough <- c(
  1.72, 2.50, 2.16, 2.13, 1.06, 2.24, 2.31, 2.03, 1.09, 1.40,
  2.57, 2.64, 1.26, 2.05, 1.19, 2.13, 1.27, 1.51, 2.41, 1.95
)

n  <- length(rough)
xbar <- mean(rough)
s <- sd(rough)

# 1) Empirical Rule (Normal-ish): x̄ ± 2s
range_empirical <- c(xbar - 2*s, xbar + 2*s)

# 2) Nonparametric (no distribution assumption): central 95%
range_quantile <- quantile(rough, probs = c(0.025, 0.975), type = 7)

list(
  n = n,
  mean = round(xbar, 3),
  sd = round(s, 3),
  Empirical_95pct_xbar_pm_2s = round(range_empirical, 3),
  Quantile_95pct = round(range_quantile, 3)
)

```
Interpretation

Both methods suggest that about 95% of pipe roughness values fall between ~1.0 and ~2.9 micrometers.

The empirical rule assumes an approximately normal distribution.

The quantile interval makes no distribution assumption and is narrower.

# Task 13
## Data

```{r}
# Number of ant species (all 11 sites)
ants_species <- c(3, 3, 52, 7, 5, 49, 5, 4, 56, 22, 14)

# Total plant cover (%)
plant_cover <- c(40, 52, 40, 43, 27, 30, 16, 30, 56, 22, 14)

# Split into regions
dry_species <- ants_species[1:5]
gobi_species <- ants_species[6:11]

dry_cover <- plant_cover[1:5]
gobi_cover <- plant_cover[6:11]
```

## (a) Mean, Median, Mode of Number of Ant Species

```{r}
mean_species <- mean(ants_species)
median_species <- median(ants_species)
mode_species <- as.numeric(names(which.max(table(ants_species))))

list(Mean = mean_species, Median = median_species, Mode = mode_species)

```

## (b) Which measure of central tendency?

Answer: Since the data include an extreme outlier (52 species at Site 3), the median is a better measure of central tendency than the mean.

## (c) Dry Steppe Sites – Plant Cover %
```{r}
list(
  Mean = mean(dry_cover),
  Median = median(dry_cover),
  Mode = as.numeric(names(which.max(table(dry_cover))))
)

```

## (d) Gobi Desert Sites – Plant Cover %
```{r}
list(
  Mean = mean(gobi_cover),
  Median = median(gobi_cover),
  Mode = as.numeric(names(which.max(table(gobi_cover))))
)

```
## (e) Compare the Two Regions

Answer: The Dry Steppe has higher plant cover (mean and median) compared to the Gobi Desert.
This suggests the two regions have distinct vegetation patterns, which may influence the number of ant species observed.

# Task 14

Astronomers theorize that cold dark matter (CDM) caused the formation of galaxies.  
One galaxy, A1775, is thought to be a *double cluster*. Data from 51 galaxy velocities  
(in km/s) are given below.

```{r}
# Galaxy velocity data
galaxy <- c(
  22922, 20210, 21911, 19225, 18792, 21993, 23059,
  20785, 22781, 23303, 22192, 19462, 19057, 23017,
  20186, 23292, 19408, 24909, 19866, 22891, 23121,
  19673, 23261, 22796, 22355, 19807, 23432, 22625,
  22744, 22426, 19111, 18933, 22417, 19595, 23408,
  22809, 19619, 22738, 18499, 19130, 23220, 22647,
  22718, 22779, 19026, 22513, 19740, 22682, 19179,
  19404, 22193
)

length(galaxy)

```

## (a) Graphical Method

```{r}

hist(galaxy, breaks=12, col="lightblue", main="Histogram of Galaxy Velocities (A1775)",
     xlab="Velocity (km/s)")
boxplot(galaxy, horizontal=TRUE, col="lightgreen", main="Boxplot of Galaxy Velocities")
```

## (b) Evidence for Double Cluster?

From the histogram/boxplot, we expect to see two peaks (bimodal distribution),
which supports the theory of two clusters (A1775A and A1775B).

## (c) Numerical Descriptive Measures

```{r}

# Overall descriptives
overall <- c(
  n = length(galaxy),
  mean = mean(galaxy),
  sd = sd(galaxy),
  median = median(galaxy),
  min = min(galaxy),
  max = max(galaxy)
)
overall

# ---- Discover two clusters automatically (no arbitrary cut) ----
set.seed(42)
km <- kmeans(matrix(galaxy, ncol = 1), centers = 2, nstart = 50)

# Order cluster labels by their means so: A = lower-mean, B = higher-mean
cl_means <- tapply(galaxy, km$cluster, mean)
ord <- order(cl_means)                # index of lower, higher
# Relabel to A/B
clusterAB <- ifelse(km$cluster == which(ord == 1), "A1775A", "A1775B")

# Descriptives by cluster
by_mean <- tapply(galaxy, clusterAB, mean)
by_sd   <- tapply(galaxy, clusterAB, sd)
by_n    <- tapply(galaxy, clusterAB, length)
by_sum  <- rbind(n = by_n, mean = by_mean, sd = by_sd)
round(by_sum, 2)


```

#Visual check

```{r}

hist(galaxy, breaks = 12, col = "lightgray",
     main = "Galaxy Velocities (A1775) with Cluster Means",
     xlab = "Velocity (km/s)")
abline(v = by_mean["A1775A"], col = "blue",  lwd = 2, lty = 2)
abline(v = by_mean["A1775B"], col = "red",   lwd = 2, lty = 2)
legend("topleft",
       legend = c(paste0("Mean A: ", round(by_mean["A1775A"],0)),
                  paste0("Mean B: ", round(by_mean["A1775B"],0))),
       col = c("blue","red"), lwd = 2, lty = 2, bty = "n")
```

## (d) Probability of 20,000 km/s Galaxy


```{r}
x0 <- 20000

muA <- by_mean["A1775A"]; sdA <- by_sd["A1775A"]; nA <- by_n["A1775A"]
muB <- by_mean["A1775B"]; sdB <- by_sd["A1775B"]; nB <- by_n["A1775B"]
piA <- nA / (nA + nB);     piB <- nB / (nA + nB)

# 1) Z-scores
zA <- (x0 - muA)/sdA
zB <- (x0 - muB)/sdB

# 2) Posterior via normal densities with mixing proportions
likA <- piA * dnorm(x0, mean = muA, sd = sdA)
likB <- piB * dnorm(x0, mean = muB, sd = sdB)
postA <- likA / (likA + likB)
postB <- 1 - postA

list(
  z_in_A = round(zA, 2),
  z_in_B = round(zB, 2),
  posterior_P_A = round(postA, 3),
  posterior_P_B = round(postB, 3)
)

best_cluster <- if (postA >= postB) "A1775A" else "A1775B"
best_cluster


```


# Task 15
```{r, message=FALSE, warning=FALSE}
# Load required packages
library(ggplot2)

# Read the dataset (adjust the path if needed)
ddt <- read.csv("DDT.csv", header = TRUE)

# Preview the structure of the data
head(ddt)

# Create the boxplot
ggplot(ddt, aes(x=RIVER, y=LENGTH, fill=SPECIES)) +
  geom_boxplot() +
  scale_fill_manual(values = c("CCATFISH" = "red",
                               "LMBASS"   = "green",
                               "SMBUFFALO"= "blue")) +
  labs(title = "Fish Lengths by River and Species - Nishant Ahlawat",
       x = "RIVER",
       y = "LENGTH",
       fill = "SPECIES") +
  theme_minimal(base_size = 14)

```
